Project Report on DeepSpecDB

### 2018-08-28 papers/luke/report.tex
turn table into map

Within the index engine, MassTree is naturally divided into two modules: A
module that implements B+ tree with cursors and a module that relies on previous
module to implement trie with cursors. In the verification of trie, we went a
further step where we parameterized over the dependency: for any module that
implements the specification of table of integer-typed key with
cursors, we can derive a module that implements the same specification of
string-typed key with cursors using the trie data structure. Such
parameterization modularized the verification in the way that verification of
trie is now independent from any detail in implementation or verification of B+
trees. Also as a side effect, this approach \textbf{tested} the
\textit{two-sidedness} of the specification: As a consumer of the specification,
a successful verification implied that the specification was sufficient in
capturing the behaviors. Meanwhile as a provider, it made sure that only
reasonable properties was specified.

@@

\section{Specification for Table with Cursors}\label{spec}

In this section I will introduce the specification for indexing data
structures. The specification can be generally divided into two parts: the part
where it guaranteed the correctness of the table itself and the part where it
provided essential utilities for clients' reasoning.

@@

Spiritually, a table maintains the mapping from keys to values and a cursor
points into the table at a specific position and supports movement operations.
Concretely, such a data structure should provide the operations in
figure~\ref{fig:interface}.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  Definition key := KeyType.t.
  Parameter table: Type -> Type.
  Parameter cursor: Type -> Type.

  Section Types.
    Context {value: Type}.
    Parameter empty: table value -> Prop.

    Parameter make_cursor: key -> table value -> cursor value.
    Parameter first_cursor: table value -> cursor value.
    Parameter last_cursor: table value -> cursor value.
    Parameter next_cursor: cursor value -> table value -> cursor value.
    Parameter prev_cursor: cursor value -> table value -> cursor value.

    Parameter get: cursor value -> table value -> option (key * value).

    Parameter put: key -> value -> cursor value -> table value -> (* input *)
                   cursor value -> table value -> (* output *)
                   Prop.
    [...]
  End Types.
End ABSTRACT_TABLE.
\end{minted}
  \caption{Interface for Table with Cursors}\label{fig:interface}
\end{figure}

This is a typical interface of tables which supports operations for creation of
table (\texttt{empty}), creation of cursor (\texttt{make\_cursor},
\texttt{first\_cursor}, \texttt{last\_cursor}), movement of cursor
(\texttt{next\_cursor}, \texttt{prev\_cursor}), fetching data at
cursor (\texttt{get}) and updating the table (\texttt{put}). The types of
functions indicates clearly that all other functions except for \texttt{empty}
and \texttt{put} are immutable with respect to the table. The two table mutating
functions are defined as relations rather than functions, this is a compromise
to support cursors. I will discuss the reason and consequence of relational
specification in detail later in this section.

Apart from these functions, we also need some auxiliary predicates, as shown in
figure~\ref{fig:aux}. \texttt{abs\_rel} asserts that a cursor and a table are
correctly associated with each other (the abbreviation \texttt{abs} should
stands for \emph{abstraction}, but it's not a right word here, rather a
historical issue). \texttt{key\_rel} asserts that a cursor is associated with
a key in a table. One can imagine this predicate as \textit{the cursor is placed
at the position such that all the existing keys to the decreasing side are less
than this key, and all the existing keys to the opposite are no less than this
key}. This makes sense when one visualizes a table as a list of pairs of key and
value. This definition is closely related to the semantics of \texttt{get}, We
will see later how this is used in generalized tables. \texttt{eq\_cursor}
asserts that two cursors are equivalent. Coq equality (\texttt{eq}) does not
work well here, because of \texttt{eq} implies structural equality, and
structurally different cursors can be equivalent (especially for those table
with complex structures, such as B+ trees or tries). The final two predicates
\texttt{cursor\_correct} and \texttt{table\_correct} are the normal invariants.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Parameter abs_rel: cursor value -> table value -> Prop.
  Parameter key_rel: key -> cursor value -> table value -> Prop.
  Parameter eq_cursor : cursor value -> cursor value -> table value -> Prop.
  Parameter cursor_correct: cursor value -> Prop.
  Parameter table_correct: table value -> Prop.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Auxiliary Predicates of Table with Cursors}\label{fig:aux}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom empty_correct: forall t, empty t -> table_correct t.
  Axiom get_empty: forall t c, empty t -> abs_rel c t -> get c t = None.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{empty}}\label{fig:empty}
\end{figure}

For an empty table, it should satisfy the correctness invariant and also that we
are never able to get anything from cursor associated with the table.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom make_cursor_key: forall t k,
    table_correct t -> key_rel k (make_cursor k t) t.
  Axiom make_cursor_abs: forall t k,
    table_correct t -> abs_rel (make_cursor k t) t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{make\_cursor}}\label{fig:make}
\end{figure}

For the \texttt{make\_cursor} operation, the cursor it made should always be
associated with the key and the table.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom first_cursor_abs: forall t,
    table_correct t -> abs_rel (first_cursor t) t.
  Axiom last_cursor_abs: forall t,
    table_correct t -> abs_rel (last_cursor t) t.
  Axiom get_last: forall t, get (last_cursor t) t = None.
  Axiom first_cursor_get_empty: forall t,
    table_correct t -> get (first_cursor t) t = None -> empty t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{first\_cursor} and \texttt{last\_cursor}}\label{fig:first_last}
\end{figure}

\texttt{first\_cursor} and \texttt{last\_cursor} are dual operations. Both of
them should yield cursors associated with the table. Additionally, the
\texttt{last\_cursor} should point past the end of the table, meaning that we
can get nothing from it. The \texttt{first\_cursor} should always return a
binding, unless the table is empty.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom next_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (next_cursor c t) t.
  Axiom prev_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (prev_cursor c t) t.
  Axiom next_order: forall t c k1 k2,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    KeyType.lt k1 k2.
  Axiom prev_order: forall t c k1 k2,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    KeyType.lt k2 k1.
  Axiom next_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (next_cursor c t) t ->
    KeyType.lt k1 k3 /\ KeyType.lt k3 k2 ->
    False.
  Axiom prev_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (prev_cursor c t) t ->
    KeyType.lt k3 k1 /\ KeyType.lt k2 k3 ->
    False.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{next\_cursor} and \texttt{prev\_cursor}}\label{fig:next_prev}
\end{figure}

\texttt{next\_cursor} and \texttt{prev\_cursor} is another pair of dual
operations. Both of them should yield cursors associated with the table. Unless
at the end of table, movement should reach a cursor associated with keys of the
right order, they should be inverse to each other and they should move only to
adjacent positions.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom get_key_rel: forall t c k,
    abs_rel c t -> get_key c t = Some k -> key_rel k c t.

  Axiom put_correct: forall t1 t2 c1 c2 k v,
    abs_rel c1 t1 ->
    put k v c1 t1 c2 t2 ->
    table_correct t2.

  Axiom get_put_same: forall t1 t2 c1 c2 c3 k v,
    put k v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c3 t2 ->
    key_rel k c3 t2 ->
    get c3 t2 = Some (k, v).
  Axiom get_put_diff: forall t1 t2 c1 c2 c3 c4 k1 k2 v,
    k1 <> k2 ->
    put k1 v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c4 t1 ->
    key_rel k2 c4 t1 ->
    abs_rel c3 t2 ->
    key_rel k2 c3 t2 ->
    get c3 t2 = get c4 t1.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Specification for \texttt{get} and \texttt{put}}\label{fig:get_put}
\end{figure}

\texttt{get} operation are different from the \texttt{get} operation of finite
map library in the Coq standard library: \texttt{get} operation here should
fetch the first key to the increasing side of the cursor (In other word, the
largest key that is still associated with this cursor by \texttt{key\_rel}).
However, \texttt{get} and \texttt{put} should satisfy the normal permute and
shadow property as well. Additionally, \texttt{put} operation should keep a
correct table correct after mutation. Notice that, the \texttt{put} takes an
extra cursor as input but is not specified in the axioms. This cursor is there
so that some implementation might optimize the operation based on a nearby
cursor (for example, the B+ tree implemented by Aur\`ele and Oluwatosin), but it
will never affect the semantics of the operation.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type ABSTRACT_TABLE (KeyType: UsualOrderedType).
  [...]
  Axiom abs_correct: forall t c,
    abs_rel c t -> cursor_correct c /\ table_correct t.

@@

  Axiom eq_cursor_get: forall t c1 c2,
    abs_rel c1 t ->
    abs_rel c2 t ->
    eq_cursor c1 c2 t ->
    get c1 t = get c2 t.
  [...]
End ABSTRACT_TABLE.
\end{minted}
  \caption{Additional Specification}\label{fig:aux_spec}
\end{figure}

In addition to the semantics of operations, more specifications are there to
regulate the predicates. \texttt{abs\_rel} should imply that both the cursor and
table correct. Cursors associated with the same key should be equivalent and
equivalent cursors should return the same result.

@@

However, such types will introduce another issue: The pointers are only given at
runtime from the \texttt{malloc} calls. They do not appear in the inputs and can
vary in different runs. The nondeterminism cannot be captured well if the
interface is defined with functions. Therefore for those functions that mutate
the table, we are compelled to define them relationally.

@@

Previous works on verification of B+ tree had assumed that the \texttt{value}
type are integers (\texttt{Int.int}). However, clients usually want a map over
arbitrary \texttt{value} types. Some clients even want to define inductive types
with the table. A minimal example is the finite-degree tree where at each node
it maintains a map from keys to children. However, such definition will never
pass the positiveness check, as a consequence of abstraction.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Tree (K: UsualOrderedType) (NodeTable: Table K).
  Inductive tree: Type :=
  | node_of: NodeTable.table tree -> tree.
End Tree.
\end{minted}
  \caption{Finite-degree Tree}\label{fig:inductive}
\end{figure}

Even if the client does not require strict positiveness, a separation logic
client might need to quantify over all the bindings and assert that they are
well represented in the memory, as shown in Figure~\ref{fig:rep}. There have
been quite a few new functions declared without specification, and even more are
required to work with table mutations. As the amount of new functions suggest,
additional specification should be added so that separation logic clients live a
better life.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Client (K: UsualOrderedType) (T: Table K).
  Parameter value: Type.
  Parameter value_rep: value -> val -> mpred.
  Definition t: Type := T.table value.

  Definition rep: (data: t) (p: val): mpred :=
    EX concrete_table: T.table val,
    !! (Table.keys concrete_table = Table.keys data) &&
    Table.table_rep concrete_table p *
    Table.iter_sepcon data concrete_table value_rep.
End Tree.
\end{minted}
  \caption{A Painful Separation Logic Client}\label{fig:rep}
\end{figure}

The remedy for both issues is the \texttt{flatten} function. \texttt{flatten}
turns a table into list of pairs of key and value, where the list is sorted in
the order of keys. Furthermore, The list have been proven to be an
implementation of the same interface as table with cursors. Formally, these
properties are specified in Figure~\ref{fig:flatten}.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FLATTENABLE_TABLE (KeyType: UsualOrderedType)
                              <: ABSTRACT_TABLE KeyType.
  Include ABSTRACT_TABLE KeyType.
  Module Flattened := SortedListTable KeyType.
  Section Spec.
    Context {value: Type}.
    Parameter flatten: table value -> Flattened.table value.
    Axiom flatten_invariant: forall t,
        table_correct t ->
        Flattened.table_correct (flatten t) /\
        forall (k: key) (c1: cursor value) (c2: Flattened.cursor value),
          key_rel k c1 t -> Flattened.key_rel k c2 (flatten t) ->
          abs_rel c1 t -> Flattened.abs_rel c2 (flatten t) ->
          get c1 t = Flattened.get c2 (flatten t).
  End Spec.
End FLATTENABLE_TABLE.
\end{minted}
  \caption{Flattenable Table}\label{fig:flatten}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FlattenableTableFacts (KeyType: UsualOrderedType)
                                  (Table: FLATTENABLE_TABLE KeyType).
  Include Table.
  Section Implication.
    Context {value: Type}.
    Theorem put_permute: forall t1 t2 c1 c2 k v,
      table_correct t1 ->
      abs_rel c1 t1 ->
      put k v c1 t1 c2 t2 ->
      flatten t2 = Flattened.fput k v (flatten t1).

    Theorem Forall_put (P: value -> Prop): forall t k v,
      Flattened.table_correct t ->
      P v ->
      Forall P (t) ->
      Forall P (Flattened.fput k v t).
    Theorem iter_sepcon_put (P: value -> mpred): forall t k v,
      Flattened.table_correct t ->
      P v * iter_sepcon P (flatten t) |--
      (option_wrap P (Flattened.get_exact k t)) *
      iter_sepcon P (Flattened.fput k v t).
End FLATTENABLE_TABLE.
\end{minted}
  \caption{Facts about Flattenable Table}\label{fig:flatten_fact}
\end{figure}

@@

Figure~\ref{fig:typedef} defines the type for functional trie. A trie is defined
inductively, containing an address, a node table maps from \texttt{keyslice}s to
addresses and a flattened node table maps from \texttt{keyslice}s to a pair of
address and \texttt{BorderNode}. While \texttt{BorderNode} can optionally points
to next level of trie or final value, as represented by the \texttt{link} type.
One should have noticed that there is a duplication of information: both the
table form of node and list form of node is saved. We need the table form
because of we need the unmodified augmented types for representation in
separation logic. We need the list form to pass the positiveness check. This is
the original motivation to introduce \texttt{flatten}. A \texttt{BorderNode}
contains a list of prefix links, an optional suffix key field and a suffix link
field. When there is a suffix key, the suffix link field should point to a
final value. Otherwise, the suffix link field should optionally point to
another trie. Although the \texttt{BorderNode}'s ``value'' type is always the
same. Different endpoints (prefix links and suffix link) have different
requirement and not all three constructors are valid, future work might refine
this part better.

@@

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module BorderNode.
  [...]
  Definition table: Type := list value * option string * value.
  Inductive cursor: Type :=
    | before_prefix: Z -> cursor
    | before_suffix: cursor
    | after_suffix: cursor.
  [...]
End BorderNode.

@@

Module Trie (Node: FLATTENABLE_TABLE Keyslice) <: FLATTENABLE_TABLE KeyType.
  [...]
  Inductive trie: Type :=
  | trienode_of: val ->
                 Node.table val ->
                 Node.Flattened.table (val * BorderNode.table link) ->
                 trie
  with link: Type :=
  | value_of: value -> link
  | trie_of: trie -> link
  | nil: link.

  Definition cursor: Type := list
     (trie * Node.cursor val * (BorderNode.table link) * BorderNode.cursor).
  [...]
End Trie.
\end{minted}
  \caption{Trie Definition}\label{fig:typedef}
\end{figure}

@@

The \texttt{empty} predicate is defined straightforward. The root node of the trie
should contain only empty node table. There exists another possible definition
where we introduce a \texttt{empty} constructor for \texttt{trie} type, however.
Which one is better is still under question.

@@

\texttt{first\_cursor} is a wrapper of \texttt{strict\_first\_cursor}. If the
auxiliary function failed to find a cursor, then \texttt{first\_cursor} will
return an empty cursor, which is regarded as \texttt{last\_cursor} for any
tables.

@@

\subsubsection{\texttt{table\_correct}}

For a table to be correct, the following properties should hold:

>>>>

Within the index engine, MassTree is naturally divided into two modules: A
module that implements B+ tree with cursors and a module that relies on previous
module to implement trie with cursors. In the verification of trie, we went a
further step where we parameterized over the dependency: for any module that
implements the specification of kv-map of integer-typed key with
cursors, we can derive a module that implements the same specification of
string-typed key with cursors using the trie data structure. Such
parameterization modularized the verification in the way that verification of
trie is now independent from any detail in implementation or verification of B+
trees. Also as a side effect, this approach \textbf{tested} the
\textit{two-sidedness} of the specification: As a consumer of the specification,
a successful verification implied that the specification was sufficient in
capturing the behaviors. Meanwhile as a provider, it made sure that only
reasonable properties was specified.

@@

\section{Specification for KV-Map with Cursors}\label{spec}

In this section I will introduce the specification for indexing data
structures. The specification can be generally divided into two parts: the part
where it guaranteed the correctness of the kv-map itself and the part where it
provided essential utilities for clients' reasoning.

@@

Spiritually, a kv-map maintains the mapping from keys to values and a cursor
points into the kv-map at a specific position and supports movement operations.
Concretely, such a data structure should provide the operations in
figure~\ref{fig:interface}.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  Definition key := KeyType.t.
  Parameter map: Type -> Type.
  Parameter cursor: Type -> Type.

  Section Types.
    Context {value: Type}.
    Parameter empty: map value -> Prop.

    Parameter make_cursor: key -> map value -> cursor value.
    Parameter first_cursor: map value -> cursor value.
    Parameter last_cursor: map value -> cursor value.
    Parameter next_cursor: cursor value -> map value -> cursor value.
    Parameter prev_cursor: cursor value -> map value -> cursor value.

    Parameter get: cursor value -> map value -> option (key * value).

    Parameter put: key -> value -> cursor value -> map value -> (* input *)
                   cursor value -> map value -> (* output *)
                   Prop.
    [...]
  End Types.
End KV_MAP.
\end{minted}
  \caption{Interface for KV-Map with Cursors}\label{fig:interface}
\end{figure}

This is a typical interface of kv-map which supports operations for creation of
map (\texttt{empty}), creation of cursor (\texttt{make\_cursor},
\texttt{first\_cursor}, \texttt{last\_cursor}), movement of cursor
(\texttt{next\_cursor}, \texttt{prev\_cursor}), fetching data at
cursor (\texttt{get}) and updating the map (\texttt{put}). The types of
functions indicates clearly that all other functions except for \texttt{empty}
and \texttt{put} are immutable with respect to the map. The two map mutating
functions are defined as relations rather than functions, this is a compromise
to support cursors. I will discuss the reason and consequence of relational
specification in detail later in this section.

Apart from these functions, we also need some auxiliary predicates, as shown in
figure~\ref{fig:aux}. \texttt{abs\_rel} asserts that a cursor and a kv-map are
correctly associated with each other (the abbreviation \texttt{abs} should
stands for \emph{abstraction}, but it's not a right word here, rather a
historical issue). \texttt{key\_rel} asserts that a cursor is associated with
a key in a kv-map. One can imagine this predicate as \textit{the cursor is placed
at the position such that all the existing keys to the decreasing side are less
than this key, and all the existing keys to the opposite are no less than this
key}. This makes sense when one visualizes a kv-map as a list of pairs of key and
value. This definition is closely related to the semantics of \texttt{get}, We
will see later how this is used in generalized kv-map. \texttt{eq\_cursor}
asserts that two cursors are equivalent. Coq equality (\texttt{eq}) does not
work well here, because of \texttt{eq} implies structural equality, and
structurally different cursors can be equivalent (especially for those kv-map
with complex structures, such as B+ trees or tries). The final two predicates
\texttt{cursor\_correct} and \texttt{map\_correct} are the normal invariants.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Parameter abs_rel: cursor value -> map value -> Prop.
  Parameter key_rel: key -> cursor value -> map value -> Prop.
  Parameter eq_cursor : cursor value -> cursor value -> map value -> Prop.
  Parameter cursor_correct: cursor value -> Prop.
  Parameter map_correct: map value -> Prop.
  [...]
End KV_MAP.
\end{minted}
  \caption{Auxiliary Predicates of KV-Map with Cursors}\label{fig:aux}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom empty_correct: forall t, empty t -> map_correct t.
  Axiom get_empty: forall t c, empty t -> abs_rel c t -> get c t = None.
  [...]
End KV_MAP.
\end{minted}
  \caption{Specification for \texttt{empty}}\label{fig:empty}
\end{figure}

For an empty map, it should satisfy the correctness invariant and also that we
are never able to get anything from cursor associated with the kv-map.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom make_cursor_key: forall t k,
    map_correct t -> key_rel k (make_cursor k t) t.
  Axiom make_cursor_abs: forall t k,
    map_correct t -> abs_rel (make_cursor k t) t.
  [...]
End KV_MAP.
\end{minted}
  \caption{Specification for \texttt{make\_cursor}}\label{fig:make}
\end{figure}

For the \texttt{make\_cursor} operation, the cursor it made should always be
associated with the key and the kv-map.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom first_cursor_abs: forall t,
    map_correct t -> abs_rel (first_cursor t) t.
  Axiom last_cursor_abs: forall t,
    map_correct t -> abs_rel (last_cursor t) t.
  Axiom get_last: forall t, get (last_cursor t) t = None.
  Axiom first_cursor_get_empty: forall t,
    map_correct t -> get (first_cursor t) t = None -> empty t.
  [...]
End KV_MAP.
\end{minted}
  \caption{Specification for \texttt{first\_cursor} and \texttt{last\_cursor}}\label{fig:first_last}
\end{figure}

\texttt{first\_cursor} and \texttt{last\_cursor} are dual operations. Both of
them should yield cursors associated with the kv-map. Additionally, the
\texttt{last\_cursor} should point past the end of the kv-map, meaning that we
can get nothing from it. The \texttt{first\_cursor} should always return a
binding, unless the map is empty.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom next_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (next_cursor c t) t.
  Axiom prev_cursor_abs: forall t c,
    abs_rel c t -> abs_rel (prev_cursor c t) t.
  Axiom next_order: forall t c k1 k2,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    KeyType.lt k1 k2.
  Axiom prev_order: forall t c k1 k2,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    KeyType.lt k2 k1.
  Axiom next_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (last_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (next_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (next_cursor c t) t ->
    KeyType.lt k1 k3 /\ KeyType.lt k3 k2 ->
    False.
  Axiom prev_compat: forall t c k1 k2 k3,
    ~ eq_cursor c (first_cursor t) t ->
    key_rel k1 c t -> key_rel k2 (prev_cursor c t) t ->
    ~ key_rel k3 c t -> ~ key_rel k3 (prev_cursor c t) t ->
    KeyType.lt k3 k1 /\ KeyType.lt k2 k3 ->
    False.
  [...]
End KV_MAP.
\end{minted}
  \caption{Specification for \texttt{next\_cursor} and \texttt{prev\_cursor}}\label{fig:next_prev}
\end{figure}

\texttt{next\_cursor} and \texttt{prev\_cursor} is another pair of dual
operations. Both of them should yield cursors associated with the kv-map. Unless
at the end of kv-map, movement should reach a cursor associated with keys of the
right order, they should be inverse to each other and they should move only to
adjacent positions.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom get_key_rel: forall t c k,
    abs_rel c t -> get_key c t = Some k -> key_rel k c t.

  Axiom put_correct: forall t1 t2 c1 c2 k v,
    abs_rel c1 t1 ->
    put k v c1 t1 c2 t2 ->
    map_correct t2.

  Axiom get_put_same: forall t1 t2 c1 c2 c3 k v,
    put k v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c3 t2 ->
    key_rel k c3 t2 ->
    get c3 t2 = Some (k, v).
  Axiom get_put_diff: forall t1 t2 c1 c2 c3 c4 k1 k2 v,
    k1 <> k2 ->
    put k1 v c1 t1 c2 t2 ->
    abs_rel c1 t1 ->
    abs_rel c4 t1 ->
    key_rel k2 c4 t1 ->
    abs_rel c3 t2 ->
    key_rel k2 c3 t2 ->
    get c3 t2 = get c4 t1.
  [...]
End KV_MAP.
\end{minted}
  \caption{Specification for \texttt{get} and \texttt{put}}\label{fig:get_put}
\end{figure}

\texttt{get} operation are different from the \texttt{get} operation of finite
map library in the Coq standard library: \texttt{get} operation here should
fetch the first key to the increasing side of the cursor (In other word, the
largest key that is still associated with this cursor by \texttt{key\_rel}).
However, \texttt{get} and \texttt{put} should satisfy the normal permute and
shadow property as well. Additionally, \texttt{put} operation should keep a
correct kv-map correct after mutation. Notice that, the \texttt{put} takes an
extra cursor as input but is not specified in the axioms. This cursor is there
so that some implementation might optimize the operation based on a nearby
cursor (for example, the B+ tree implemented by Aur\`ele and Oluwatosin), but it
will never affect the semantics of the operation.

\begin{figure}[htbp]
  \centering
  \begin{minted}{coq}
Module Type KV_MAP (KeyType: UsualOrderedType).
  [...]
  Axiom abs_correct: forall t c,
    abs_rel c t -> cursor_correct c /\ map_correct t.

@@

  Axiom eq_cursor_get: forall t c1 c2,
    abs_rel c1 t ->
    abs_rel c2 t ->
    eq_cursor c1 c2 t ->
    get c1 t = get c2 t.
  [...]
End KV_MAP.
\end{minted}
  \caption{Additional Specification}\label{fig:aux_spec}
\end{figure}

In addition to the semantics of operations, more specifications are there to
regulate the predicates. \texttt{abs\_rel} should imply that both the cursor and
map correct. Cursors associated with the same key should be equivalent and
equivalent cursors should return the same result.

@@

However, such types will introduce another issue: The pointers are only given at
runtime from the \texttt{malloc} calls. They do not appear in the inputs and can
vary in different runs. The nondeterminism cannot be captured well if the
interface is defined with functions. Therefore for those functions that mutate
the kv-map, we are compelled to define them relationally.

@@

Previous works on verification of B+ tree had assumed that the \texttt{value}
type are integers (\texttt{Int.int}). However, clients usually want a map over
arbitrary \texttt{value} types. Some clients even want to define inductive types
with the kv-map. A minimal example is the finite-degree tree where at each node
it maintains a map from keys to children. However, such definition will never
pass the positiveness check, as a consequence of abstraction.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Tree (K: UsualOrderedType) (NodeMap: KV_MAP K).
  Inductive tree: Type :=
  | node_of: NodeMap.map tree -> tree.
End Tree.
\end{minted}
  \caption{Finite-degree Tree}\label{fig:inductive}
\end{figure}

Even if the client does not require strict positiveness, a separation logic
client might need to quantify over all the bindings and assert that they are
well represented in the memory, as shown in Figure~\ref{fig:rep}. There have
been quite a few new functions declared without specification, and even more are
required to work with map mutations. As the amount of new functions suggest,
additional specification should be added so that separation logic clients live a
better life.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Client (K: UsualOrderedType) (T: KV_MAP K).
  Parameter value: Type.
  Parameter value_rep: value -> val -> mpred.
  Definition t: Type := T.map value.

  Definition rep: (data: t) (p: val): mpred :=
    EX concrete_map: T.map val,
    !! (T.keys concrete_map = T.keys data) &&
    T.map_rep concrete_map p *
    T.iter_sepcon data concrete_map value_rep.
End Tree.
\end{minted}
  \caption{A Painful Separation Logic Client}\label{fig:rep}
\end{figure}

The remedy for both issues is the \texttt{flatten} function. \texttt{flatten}
turns a map into list of pairs of key and value, where the list is sorted in
the order of keys. Furthermore, The list have been proven to be an
implementation of the same interface as map with cursors. Formally, these
properties are specified in Figure~\ref{fig:flatten}.

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FLATTENABLE_MAP (KeyType: UsualOrderedType)
                              <: KV_MAP KeyType.
  Include KV_MAP KeyType.
  Module Flattened := SortedListMap KeyType.
  Section Spec.
    Context {value: Type}.
    Parameter flatten: map value -> Flattened.map value.
    Axiom flatten_invariant: forall t,
        map_correct t ->
        Flattened.map_correct (flatten t) /\
        forall (k: key) (c1: cursor value) (c2: Flattened.cursor value),
          key_rel k c1 t -> Flattened.key_rel k c2 (flatten t) ->
          abs_rel c1 t -> Flattened.abs_rel c2 (flatten t) ->
          get c1 t = Flattened.get c2 (flatten t).
  End Spec.
End FLATTENABLE_MAP.
\end{minted}
  \caption{Flattenable Map}\label{fig:flatten}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module Type FlattenableMapFacts (KeyType: UsualOrderedType)
                                (Map: FLATTENABLE_MAP KeyType).
  Include Map.
  Section Implication.
    Context {value: Type}.
    Theorem put_permute: forall t1 t2 c1 c2 k v,
      map_correct t1 ->
      abs_rel c1 t1 ->
      put k v c1 t1 c2 t2 ->
      flatten t2 = Flattened.fput k v (flatten t1).

    Theorem Forall_put (P: value -> Prop): forall t k v,
      Flattened.map_correct t ->
      P v ->
      Forall P (t) ->
      Forall P (Flattened.fput k v t).
    Theorem iter_sepcon_put (P: value -> mpred): forall t k v,
      Flattened.map_correct t ->
      P v * iter_sepcon P (flatten t) |--
      (option_wrap P (Flattened.get_exact k t)) *
      iter_sepcon P (Flattened.fput k v t).
End FLATTENABLE_MAP.
\end{minted}
  \caption{Facts about Flattenable Map}\label{fig:flatten_fact}
\end{figure}

@@

Figure~\ref{fig:typedef} defines the type for functional trie. A trie is defined
inductively, containing an address, a node maps from \texttt{keyslice}s to
addresses and a flattened node maps from \texttt{keyslice}s to a pair of
address and \texttt{BorderNode}. While \texttt{BorderNode} can optionally points
to next level of trie or final value, as represented by the \texttt{link} type.
One should have noticed that there is a duplication of information: both the
map form of node and list form of node is saved. We need the map form
because of we need the unmodified augmented types for representation in
separation logic. We need the list form to pass the positiveness check. This is
the original motivation to introduce \texttt{flatten}. A \texttt{BorderNode}
contains a list of prefix links, an optional suffix key field and a suffix link
field. When there is a suffix key, the suffix link field should point to a final
value. Otherwise, the suffix link field should optionally point to another trie.
Although the \texttt{BorderNode}'s ``value'' type is always the same. Different
endpoints (prefix links and suffix link) have different requirement and not all
three constructors are valid, future work might refine this part better.

@@

\begin{figure}[htbp]
  \centering
\begin{minted}{coq}
Module BorderNode.
  [...]
  Definition map: Type := list value * option string * value.
  Inductive cursor: Type :=
    | before_prefix: Z -> cursor
    | before_suffix: cursor
    | after_suffix: cursor.
  [...]
End BorderNode.

@@

Module Trie (Node: FLATTENABLE_MAP Keyslice) <: FLATTENABLE_MAP KeyType.
  [...]
  Inductive trie: Type :=
  | trienode_of: val ->
                 Node.map val ->
                 Node.Flattened.map (val * BorderNode.map link) ->
                 trie
  with link: Type :=
  | value_of: value -> link
  | trie_of: trie -> link
  | nil: link.

  Definition cursor: Type := list
     (trie * Node.cursor val * (BorderNode.map link) * BorderNode.cursor).
  [...]
End Trie.
\end{minted}
  \caption{Trie Definition}\label{fig:typedef}
\end{figure}

@@

The \texttt{empty} predicate is defined straightforward. The root node of the trie
should contain only empty node. There exists another possible definition where
we introduce a \texttt{empty} constructor for \texttt{trie} type, however. Which
one is better is still under question.

@@

\texttt{first\_cursor} is a wrapper of \texttt{strict\_first\_cursor}. If the
auxiliary function failed to find a cursor, then \texttt{first\_cursor} will
return an empty cursor, which is regarded as \texttt{last\_cursor} for any
maps.

@@

\subsubsection{\texttt{map\_correct}}

For a trie to be correct, the following properties should hold:
